---
title: "The simplest model of logistic regression using tabulated subset data"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Model description
This simplest model describes only presence or absence of whitefish larvae $y$, not its density. Here the observational model is described as Bernoulli distribution given as: 

$$y_i |m_i \sim  Bernoulli(g^{-1}(m_i(x_i))),$$ 

where we assume logit transformation of the mean function $m(x)$

$$g(m_i(x_i))= logit(m_i(x_i)) = log\left(\frac{m_i(x_i)}{1-m_i(x_i)}\right)
$$

The mean function $m(x)$ follows the logistic regression model: 

 $$m(x) = logit^{-1}(\alpha_0 + \alpha_1\delta_1(x_1)+ ... + \alpha_6x_6(x_1) + x_2\beta_1 + ... + x_6\beta_5). $$
 
 In this model: 
 
 $\delta_d(x_1)=1$ if $x_1=d$ and zero otherwise, $\alpha_d \sim N(0,10)$ and $\beta_d \sim N(0,10), 0, ..., 5$

The priors are defined as: 

$$\alpha_d \sim N(0,10)$$ and 
$$\beta_d \sim N(0,10), d=0, ..., 5$$

The posterior distribution: 

$$p(\alpha_0, \alpha_1, ..., \alpha_6, \beta_1, ... ,\beta_5 | y,x) = p(\alpha_0)p(\alpha_l) ... p(\alpha_6)p(\beta_1)...p(\beta_5)\prod_{i=1}^{n}p(y_i | \alpha_0,\alpha_1, ..., \alpha_5,\beta_1, ... ,\beta_5, x_i)$$

## Visualizing the study area and two environmental variables

We start with loading the smaller subset data (in tabular format) and visualizing the study area. 
First we plot the variable Distance to 20m deep water at the sampling locations. The locations of samling sites are marked in black. 

```{r}
setwd("D:/Neli/MThesis/Ex_7_trials")

if (.Platform$OS.type != "windows") {
  if (Sys.getenv("RSTUDIO") == "1") {
    windows <- function( ... ) X11( ... )
  } else {
    windows <- dev.new
  }
}

library("rstan")
library(matrixStats)
options(mc.cores = parallel::detectCores())

library("raster")
library(Rcpp)
# Load the data
whitefish.dat = read.table("data_whitefish.txt", header=TRUE, sep="\t")
whitefish.raster = read.table("predraster_whitefish.txt", header=TRUE, sep="\t")

## The coordinate system in the data is ETRS89.  
## This is a 3D cartesian coordinate system constructed specifically to European areas. 
##  In whitefish.dat the coordinates are N_etrs89 E_etrs89
##  In whitefish.raster the coordinates are X Y                 

# First examine the data by plotting few of the raster layers 

# Visualize few environmental covariates
e <- extent(cbind(whitefish.raster$X,whitefish.raster$Y))
r <- raster(e, ncol=length(unique(whitefish.raster$X)), nrow=length(unique(whitefish.raster$Y)))

z <- rasterize(cbind(whitefish.raster$X,whitefish.raster$Y), r, whitefish.raster$DIST20M, fun=mean)
plot(z, xlim=cbind(min(whitefish.raster$X),max(whitefish.raster$X)),
     ylim=cbind(min(whitefish.raster$Y),max(whitefish.raster$Y)), main="Distance to 20m deep water")

# Plot the locations of sampling sites
points(whitefish.dat$E_etrs89,whitefish.dat$N_etrs89)


```

The last ice cover day in 2009 at sampling locations is visualized on the next plot: 

```{r}
# Plot Last ice cover day in 2009
z <- rasterize(cbind(whitefish.raster$X,whitefish.raster$Y), r, whitefish.raster$ICELAST09, fun=mean)
plot(z, xlim=cbind(min(whitefish.raster$X),max(whitefish.raster$X)),
     ylim=cbind(min(whitefish.raster$Y),max(whitefish.raster$Y)), main="Last ice cover day in 2009")
# Plot the locations of sampling sites
points(whitefish.dat$E_etrs89,whitefish.dat$N_etrs89)

```


## Performing logistic regression using Stan. 

The environmental covariates are: 

BOTTOMCLS - Bottom type classifiction, a categorical variable with  the following classes:
  0 = not shallow
  1 = open water
  2 = other
  3 = sand
  4 = sand/mud
  5 = sand/stone

DIS_SAND  - distance to sandy shore, continuous variable

FE300ME   - The average fetch (opennes/exposure) over all directions, continuous variable

ICELAST09 - The last ice cover date in winter 2009-10, continuous variable

RIVERS    - Influence of rivers (~weighted average distance to river mouths), continuous variable

SAL910WIN - Winter salinity in 2009-2010, continuous variable

The output variable is
  WHISUM - the number of whitefishes caught in sampling occasion. 

In this simplest case we model only presence or absence of whitefish, because of which we transform it to 0/1 binary variable. 

First we set up an standardize the data and prediction varables.

```{r}
# Set up data
s = as.matrix(cbind(whitefish.dat$E_etrs89,whitefish.dat$N_etrs89)) / 1000   # spatial coordinates in km
x = matrix(0,nrow=nrow(s),ncol=12)      # intercept + 6 BOTTOMCLS classes + 5 continues covariates
x[,1] = 1                             # Set the column corresponding to intercept to 1
x[whitefish.dat$BOTTOMCLS==0,2] = 1   # Set the elements corresponding to BOTTOMCLS = 0 to 1
x[whitefish.dat$BOTTOMCLS==1,3] = 1   # Set the elements corresponding to BOTTOMCLS = 1 to 1
x[whitefish.dat$BOTTOMCLS==2,4] = 1   # Set the elements corresponding to BOTTOMCLS = 2 to 1
x[whitefish.dat$BOTTOMCLS==3,5] = 1   # Set the elements corresponding to BOTTOMCLS = 3 to 1
x[whitefish.dat$BOTTOMCLS==4,6] = 1   # Set the elements corresponding to BOTTOMCLS = 4 to 1
x[whitefish.dat$BOTTOMCLS==5,7] = 1   # Set the elements corresponding to BOTTOMCLS = 5 to 1
xcont = as.matrix(cbind(whitefish.dat$DIS_SAND,
                        whitefish.dat$FE300ME,
                        whitefish.dat$ICELAST09,
                        whitefish.dat$RIVERS,
                        whitefish.dat$SAL910WIN))
stdxcont = apply(xcont, 2, sd)
mxcont = apply(xcont, 2, mean)
x[,8:12] = t( apply( t(apply(xcont,1,'-',mxcont)),1,'/',stdxcont) )    # "standardize" the continuous covariates

# End variable
#transform WHISUM to binary: 0 if 0, 1 otherwise
whitefish.dat$WHISUM[whitefish.dat$WHISUM > 0] <- 1
y = whitefish.dat$WHISUM               # number of counted fish larvae
#head(y)
#tail(y)

# Prediction variables
spred = as.matrix(cbind(whitefish.raster$X,whitefish.raster$Y)) / 1000  # spatial coordinates in km
xpred = matrix(0,nrow=nrow(spred),ncol=12)      # intercept + 6 BOTTOMCLS classes + 5 continues covariates
xpred[,1] = 1                             # Set the column corresponding to intercept to 1
xpred[whitefish.raster$BOTTOMCLS==0,2] = 1   # Set the elements corresponding to BOTTOMCLS = 0 to 1
xpred[whitefish.raster$BOTTOMCLS==1,3] = 1   # Set the elements corresponding to BOTTOMCLS = 1 to 1
xpred[whitefish.raster$BOTTOMCLS==2,4] = 1   # Set the elements corresponding to BOTTOMCLS = 2 to 1
xpred[whitefish.raster$BOTTOMCLS==3,5] = 1   # Set the elements corresponding to BOTTOMCLS = 3 to 1
xpred[whitefish.raster$BOTTOMCLS==4,6] = 1   # Set the elements corresponding to BOTTOMCLS = 4 to 1
xpred[whitefish.raster$BOTTOMCLS==5,7] = 1   # Set the elements corresponding to BOTTOMCLS = 5 to 1
xpredcont = as.matrix(cbind(whitefish.raster$DIS_SAND,
                            whitefish.raster$FE300ME,
                            whitefish.raster$ICELAST09,
                            whitefish.raster$RIVERS,
                            whitefish.raster$SAL910WIN))
xpred[,8:12] = t( apply( t(apply(xpredcont,1,'-',mxcont)),1,'/',stdxcont) )    # "standardize" the continuous covariates


```
The Stan hierarchical model of logistic regression is given in the chunk bellow. 
We have used  y ~ bernoulli_logit((x.beta)) instead y ~ bernoulli(inv_logit(x.beta)) since the function **bernoulli_logit** is numerically stable. 


```{stan output.var="Log_reg_whitefish"}
data {
  int<lower=1> N; //number of observations
  int<lower=1> Dx; //number of predictors + first column = intercept alpha_0
  matrix[N,Dx] x; //covariate matrix
  int<lower=0> y[N];
}

parameters {
  vector [Dx] beta;
}

model {
  //priors for alpha and beta coefficients 
  //all alpha and beta coefficients are involved in beta column vector
  //beta =[alpha0, alpha1, ..., alpha6,beta1,...,beta5]
  //Set week priors for intercept and all other covariates
  for (i in 1:Dx)
    beta[i] ~ normal(0, 10);
    
  //the likelihood  
  //y ~ bernoulli(inv_logit(x*beta)); - does not work; not stable numercally
  y ~ bernoulli_logit((x*beta));
}


```



Next we simulate and analyze the Stan model using MCMC simulations and statistics. We start 4 initial chains and check their mixing and convergence properties. 

```{r}
 x = x[seq(1,217,length=60),]
 y = y[seq(1,217,length=60)]

whitefish_dat <- list(Dx = ncol(x),
                      N = nrow(x),
                      x = x,
                      y = y)

#set initial values of the chains
#rnorm(#samples, mu,sigma)
set.seed(124)
#init1 <-list(beta=as.vector(rep(0,ncol(x))))
init1 <- list(beta=as.vector(rnorm(ncol(x),0,10)))
init2 <- list(beta=as.vector(rnorm(ncol(x),0,10)))
init3 <- list(beta=as.vector(rnorm(ncol(x),0,10)))
init4 <- list(beta=as.vector(rnorm(ncol(x),0,10)))
inits = list(init1,init2,init3,init4)

# Compile the STAN model
library(rstudioapi)
fit <- stan(file = "Log_reg_whitefish.stan", data = whitefish_dat, warmup=200,
            iter = 1000, init=inits , control = list(adapt_delta = 0.99), pars=c("beta"))
beta <- as.matrix(fit)

print(fit)  
#lp_ ->log density up to a constant 
# https://www.jax.org/news-and-insights/jax-blog/2015/october/lp-in-stan-output
#log density = log likelihood of the observations conditioned on the posterior parameters: p(y | p_post). 

```

We see that the potential scale reduction factor ***Rhat=1*** for all scalar beta estimates, as it should be for convergent multivariate distribution in the limit $n \rightarrow \infty$. The effective sample size ***n_eff*** is related to the precision obtained from the simulations. By one emperical rule (Gelman, Carlin, Stern, Dunson, Vehtari, and Rubin (2013). Bayesian Data Analysis, third edition. CRC Press.) $n_{eff} = 5.m$, where $m$ is twice the number of sequences, here $m=8$ and $n_{eff} = 40$. We see that here all $n_{eff} > 40$. 

A summary statistics of every chain is shown bellow:
```{r}
summary(fit) 
```

We continue to analyse MCMC convergence by visual analysis of mixing and stationarity via taceplots of all $\beta$ parameters. 

```{r}
stan_trace(fit, pars = c("beta"))
```


These plots show good mixing of all chains and therefore we can conclude that the convergence is achieved. 

Posterior densities and histograms of all estimated $\beta$ are: 
```{r}
#posterior densties and histograms
stan_hist(fit, pars='beta', bins=30)
```

The credible intervals of all $\beta$ estimates are:

```{r}
#show credible intervals
plot(fit, show_density = FALSE, ci_level = 0.5, outer_level = 0.95, fill_color = "salmon", pars='beta')

```
Pairwise correlations between estimated $\beta$ are:

```{r}
#pairwise correlations
pairs(fit)

```

Next: 

- making predictions: what thy should be? 
- visualize the predictions. 



